Cache Oblivious B-trees

### 2011-12-07 paper.tex
Adding the citations
small fixes
initial work on concurrent paper

\documentclass{acm_proc_article-sp}

@@

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Salman Ahmad\\
\affaddr{MIT CSAIL}
% 2nd. author
\alignauthor
Leilani Battle\\
\affaddr{MIT CSAIL}
% 3rd. author
\alignauthor
Stephen Tu\\
\affaddr{MIT CSAIL}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\date{7 December 2011}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

@@

@@

@@

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
%\appendix
%%Appendix A
%\section{Headings in Appendices}
%The rules about hierarchical headings discussed above for
%the body of the article are different in the appendices.
%In the \textbf{appendix} environment, the command
%\textbf{section} is used to
%indicate the start of each Appendix, with alphabetic order
%designation (i.e. the first is A, the second B, etc.) and
%a title (if you include one).  So, if you need
%hierarchical structure
%\textit{within} an Appendix, start with \textbf{subsection} as the
%highest level. Here is an outline of the body of this
%document in Appendix-appropriate form:
%\subsection{Introduction}
%\subsection{The Body of the Paper}
%\subsubsection{Type Changes and  Special Characters}
%\subsubsection{Math Equations}
%\paragraph{Inline (In-text) Equations}
%\paragraph{Display Equations}
%\subsubsection{Citations}
%\subsubsection{Tables}
%\subsubsection{Figures}
%\subsubsection{Theorem-like Constructs}
%\subsubsection*{A Caveat for the \TeX\ Expert}
%\subsection{Conclusions}
%\subsection{Acknowledgments}
%\subsection{Additional Authors}
%This section is inserted by \LaTeX; you do not insert it.
%You just add the names and information in the
%\texttt{{\char'134}additionalauthors} command at the start
%of the document.
%\subsection{References}
%Generated by bibtex from your ~.bib file.  Run latex,
%then bibtex, then latex twice (to resolve references)
%to create the ~.bbl file.  Insert that ~.bbl file into
%the .tex source file and comment out
%the command \texttt{{\char'134}thebibliography}.
%% This next section command marks the start of
%% Appendix B, and does not continue the present hierarchy
%\section{More Help for the Hardy}
%The acm\_proc\_article-sp document class file itself is chock-full of succinct
%and helpful comments.  If you consider yourself a moderately
%experienced to expert user of \LaTeX, you may find reading
%it useful but please remember not to change it.
\balancecolumns
% That's all folks!
\end{document}

>>>>

\documentclass{style}

\usepackage{paralist}

@@

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Salman Ahmad\\
\affaddr{MIT CSAIL}
\email{saahmad@mit.edu}
% 2nd. author
\alignauthor
Leilani Battle\\
\affaddr{MIT CSAIL}
\email{leibatt@mit.edu}
% 3rd. author
\alignauthor
Stephen Tu\\
\affaddr{MIT CSAIL}
\email{stephent@mit.edu}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\date{7 December 2011}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\newcommand{\lhyperceil}{\lceil\lceil}
\newcommand{\rhyperceil}{\rceil\rceil}
\newcommand{\lhyperfloor}{\lfloor\lfloor}
\newcommand{\rhyperfloor}{\rfloor\rfloor}

\newcommand{\Search}{\textsc{Search(k)}}
\newcommand{\Insert}{\textsc{Insert(k, v)}}
\newcommand{\Scan}{\textsc{Scan(a, b)}}

@@

\cite{BenderFaFi07}

@@

\subsection{Motivation}
B-trees are often used as an indexing data structure for various
database and file systems. Thus, for a cache-oblivious B-tree data structure
to be of practical value, it must be able to handle concurrent access and modification
of its internal state. The straw-man solution is to simply ensure mutual exclusion
on the entire data structure; this approach, however, yields poor performance.

Various concurrency schemes for traditional cache-aware B-trees have been
proposed \cite{BayerS77, LehmanY81}. We now survey a concurrency scheme for CO
B-trees, first presented in \cite{BenderFiGi05}. Bender develops two different
data structures to solve this problem: the first is an exponential CO B-tree,
the second is a CO B-tree based on a packed memory array (PMA). Bender
also presents a lock-free variant of the latter data structure. For
the remainder of the discussion, assume that the keys are fixed-size integer
keys.

\subsection{Description}
Before describing the data structures in more detail, we first consider
the concurrency model which is used throughout \cite{BenderFiGi05}.

\subsubsection{Concurrency Model}
The concurrency model starts with a system that has a total cache
size of $M$ and block size of $B$. There are $P$ processors in the
system, each with a cache of size $\frac{M}{P}$. A particular block
can reside in multiple caches in \textit{shared} mode. However, if a
processor wants to perform a write to a block, it must acquire it in 
\textit{exclusive} mode, in which case that block can only reside in
a single cache. In this model, a request for a block costs a single
memory transfer.

For concurrency control, two primitives are used. The first is
locks. The second is \textit{load-linked/store-conditional} (LL/SC).
The authors argue that LL/SC provides better performance than 
using a \textit{compare-and-swap} (CAS) primitive, but their 
techniques can be extended to use CAS.

\subsection{Exponential CO B-tree}
This data structure is based on a strongly weight-balanced 
exponential tree found in \cite{Bender2002}. A tuning parameter
$1 < \alpha < 2$ is chosen for the data structure. Each node
in the tree maintains both keys which divide up its children,
in addition to a \textit{right-link} pointer, which contains
both a pointer to a node's right sibling, plus the key
value of the minimum key in the node's right sibling. If no such
sibling exists, a sentinel $\infty$ pointer is used.

A node's layout in memory is as follows. Suppose a node has $k$
elements in it. The first part of the node consists of a static
CO search B-tree, containing the $k$ elements. The tree is sized
to be a complete $\lhyperceil k \rhyperceil$-leaf tree. The second
part of the node consists of a $\lhyperceil k \rhyperceil$-size array
contained the $k$ elements in sorted order. The leaf nodes of the
CO search tree point to the associated entries in the array.

\subsubsection{Operations}
The operations supported by the exponential CO B-tree are \Search{}
and \Insert{}. The authors do not discuss range-queries, and this
is presumably due to the concurrency issues in implementing a fully-consistent
range scan (weakly-consistent range scans could be implemented in the data
structure with very little modification), although interestingly enough
the PMA variant of B-tree (discussed in Section \ref{sec:pma}) does
cover range scans.

\Search{} works as follows. For a non-leaf node, first check the
\textit{right-link} key. If the \textit{right-link} key is less than
$k$, then traverse the right link, and recurse. Otherwise, follow
the appropriate child pointer. When a leaf node is reached, only
follow \textit{right-link} pointers, until either $k$ has been located
or some element greater than $k$ is reached. No locks are acquired
during the entire search operation.

\Insert{} works as follows. First, \Search{} is used to locate the leaf node
where $k$ will be inserted into. A lock is acquired on that node, $k$ is
inserted, and the lock is released. Unlike a standard B-tree, there is no
maximum number of elements in a node; instead we split nodes probabilistically.
With probability ${1}/{2^{\alpha^h}}$, a node at height $h$ is split (leaf
nodes have $h = 0$). A split works as follows.  We reacquire a lock on the node
to be split. We split the node into $u$ and $u'$, where $u$ contains all the
keys $< k$, and $u'$ contains the keys $\geq k$. We release the lock, and
recursively insert $k$ into the parent of the split node. It is important
to note that, since $u$ contains a prefix of the original node to be split,
that we reuse the original node in constructing $u$ (that way, concurrent
searches can still locate the keys in $u$).

%TODO: some section about linearizability?

\subsubsection{Complexity}

\subsection{Packed Memory Array CO B-tree}
\label{sec:pma}
The PMA CO B-tree is a two level data structure, containing a static
CO B-tree which indexes into a packed-memory array. Bender introduces
a new packed-memory array, called a \textit{one-way packed-memory array},
which they argue is easier to manage concurrently.

\subsubsection{One-Way PMA Primer}

A one-way packed-memory array is a contiguous memory structure which has three
distinct regions: a leftmost region of empty space, an \textit{active region}
which contains the elements of the array, and a rightmost region of empty
space. The array only grows into the rightmost region, and shrinks from the
leftmost region (hence one-way).  The PMA is of size $m = \Theta(N)$, subject
to $m$ being a power of $2$, where $N$ is the number of elements.

When the \textit{active region} becomes too large/small due to inserts/deletes,
the PMA is rebalanced. A new array is allocated, and the elements are
spread out evenly over the new array (too minimize element density).

Each leaf in the static CO B-tree used is a pointer into a $\Theta(\log
N)$-size region in the packed-memory array.

\subsubsection{Operations}

\Search{} works as follows. Search the static tree for $k$ until
a leaf node is reached. Follow the leaf node pointer into the PMA.
Scan right in the PMA until either $k$ is located, or some element
$> k$ is reached. Do not acquire any locks during the search.

\Insert{} works as follows. Use \Search{} to locate the $\Theta(\log N)$-region
of the PMA which (would) contain $k$. Lock the region, and scan right to find
the appropriate slot $s$ to place $k$ into. $s$ is defined such that slot $s -
1$ contains the largest key smaller than $k$. If scanning forward requires
going into the next $\Theta(\log N)$-region, lock the next region and release
the current region (\textit{hand-over-hand} locking technique). Once $s$
is located, either $s$ is free, in which case just insert $k$, or $s$ is not free,
in which case a \textit{rebalance} operation is required.

To do a rebalance operation, we scan right (locking as necessary) until the
region from $s$ to the current position $s'$ is not \textit{too dense}. The
notion of density is defined in detail in \cite{BenderFiGi05}. The region
$[s, s']$ is then rebalanced as follows. Start from the rightmost element,
and move elements to the right (maintaining order) such that no suffix of
$[s, s']$ is too dense. This rebalancing creates room for $k$; we insert $k$
and then release the locks.

\subsubsection{Complexity}

\subsubsection{Lock-free Variant}
The lock-free variant is based on the PMA CO tree, except the locks acquired
during the PMA modification are replaced by augmenting each element in the PMA
with a marker. A marker is used to denote the ongoing (concurrent) operation
of a particular cell.

In addition, four non-blocking primitives are introduced:
\begin{inparaenum}[(a)]
  \item \textit{move},
  \item \textit{cell-insertion},
  \item \textit{cell-deletion}, and
  \item \textit{read}
\end{inparaenum}. These primitives are self-explanatory, except they
all work in a non-blocking manner (they can fail at any point).

The concurrency protocol works as follows. Any of the non-blocking primitives
are initiated by denoting the cell in question with the marker. If a process
ever encounters a cell with a marker, it drops what it is doing and first helps
to complete the operation. The implementation of the non-blocking primitives
is spelled out in more detail in \cite{BenderFiGi05}, but the basic idea is
using LL/SC to detect when concurrent operations are on-going.

\subsubsection{Lock-free Variant Operations}

\Search{} is un-modified from the lock-based version.

\Insert{} is modified as follows. Cell $s$ is located as before. A
\textit{cell-insertion} operation is performed on $s$. If it succeeds, we are
done. If it fails, then a rebalance is necessary. Rebalance proceeds as before,
scanning right to find the region $[s, s']$. However, whenever a non-empty cell
is encountered, a load-link (LL) is performed on the cell.  Once the region is
found, the rebalancing-to-the-right is performed as before.  However, a
store-conditional (SC) is performed on the destination of each cell.  If any SC
fails, then this means concurrent modification had occurred, in which case the
algorithm restarts from the beginning.

@@

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{citations}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
%\appendix
%%Appendix A
%\section{Headings in Appendices}
%The rules about hierarchical headings discussed above for
%the body of the article are different in the appendices.
%In the \textbf{appendix} environment, the command
%\textbf{section} is used to
%indicate the start of each Appendix, with alphabetic order
%designation (i.e. the first is A, the second B, etc.) and
%a title (if you include one).  So, if you need
%hierarchical structure
%\textit{within} an Appendix, start with \textbf{subsection} as the
%highest level. Here is an outline of the body of this
%document in Appendix-appropriate form:
%\subsection{Introduction}
%\subsection{The Body of the Paper}
%\subsubsection{Type Changes and  Special Characters}
%\subsubsection{Math Equations}
%\paragraph{Inline (In-text) Equations}
%\paragraph{Display Equations}
%\subsubsection{Citations}
%\subsubsection{Tables}
%\subsubsection{Figures}
%\subsubsection{Theorem-like Constructs}
%\subsubsection*{A Caveat for the \TeX\ Expert}
%\subsection{Conclusions}
%\subsection{Acknowledgments}
%\subsection{Additional Authors}
%This section is inserted by \LaTeX; you do not insert it.
%You just add the names and information in the
%\texttt{{\char'134}additionalauthors} command at the start
%of the document.
%\subsection{References}
%Generated by bibtex from your ~.bib file.  Run latex,
%then bibtex, then latex twice (to resolve references)
%to create the ~.bbl file.  Insert that ~.bbl file into
%the .tex source file and comment out
%the command \texttt{{\char'134}thebibliography}.
%% This next section command marks the start of
%% Appendix B, and does not continue the present hierarchy
%\section{More Help for the Hardy}
%The acm\_proc\_article-sp document class file itself is chock-full of succinct
%and helpful comments.  If you consider yourself a moderately
%experienced to expert user of \LaTeX, you may find reading
%it useful but please remember not to change it.
\balancecolumns
% That's all folks!
\end{document}
